Re-analysing the data from Moffatt et al.~(2020): A textbook illustration of the fallacy of the null hypothesis

### 2020-09-28 reanalysis.Rmd
minor updates in manuscript and reanalysis script

---
title: "Re-analysing the data from Moffatt et al. (2020): A textbook illustration of the fallacy of the null hypothesis"
shorttitle: "The fallacy of the null"

author: 
  - name: Ladislas Nalborczyk
    affiliation: 1
    corresponding: yes
    address: GIPSA-lab, CNRS, Univ. Grenoble Alpes, 11 Rue des Mathématiques, 38400 Saint-Martin-d'Hères, France
    email: ladislas.nalborczyk@gipsa-lab.fr

affiliation:
  - id: 1
    institution: Univ. Grenoble Alpes, CNRS, Grenoble INP, GIPSA-lab, 38000 Grenoble, France

abstract: "Blah blah... Complete source code, reproducible analyses, and figures are available at https://osf.io/ntmgv/."

keywords: NHST, Bayesian, fallacy, reanalysis

csl: csl/apa7.csl
classoption: man
linenumbers: yes
floatsintext: yes
link-citations: true
numbersections: false

bibliography: [bib/brms_tutorial.bib, bib/r-references.bib]

nocite: | 
  @R-base, @R-knitr, @R-rmarkdown, @R-tidyverse, @R-papaja, @R-wordcountaddin,
  @R-here

output:
  papaja::apa6_pdf:
    keep_tex: true
    highlight: tango
    latex_engine: xelatex
---

```{r setup, message = FALSE, warning = FALSE, results = "hide"}
library(wordcountaddin)
library(rethinking)
library(extrafont)
library(rmarkdown)
library(tidyverse)
library(patchwork)
library(bayesplot)
library(ggExtra)
library(papaja)
library(knitr)
library(here)
library(brms)

@@

# setting up knitr options
knitr::opts_chunk$set(
  cache = TRUE, eval = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE,
  fig.align = "center", dev = "pdf",
  out.width = "75%", fig.pos = "!htb"
  )

@@

# Conclusions

@@

Supplementary materials, data, reproducible code and figures are available at https://osf.io/ntmgv/.

>>>>

---
title: "Re-analysing the data from Moffatt et al. (2020): A textbook illustration of the fallacy of the null hypothesis"
shorttitle: "The fallacy of the null"
author: 
  - name: Ladislas Nalborczyk
    affiliation: 1
    corresponding: yes
    address: GIPSA-lab, CNRS, Univ. Grenoble Alpes, 11 Rue des Mathématiques, 38400 Saint-Martin-d'Hères, France
    email: ladislas.nalborczyk@gipsa-lab.fr
affiliation:
  - id: 1
    institution: Univ. Grenoble Alpes, CNRS, Grenoble INP, GIPSA-lab, 38000 Grenoble, France
abstract: "Moffatt et al. (2020) reported the results of an experiment (N = 26 in the final sample) comparing the facial (surface) electromyographic correlates of mental rumination and distraction, following an experimentally induced stressful event. Based on the observation... We show ou this conclusion seems at best hasty... Indeed, concluding on the null, relation between self-reports and physiological measures... Although the question Moffatt et al. (2020) pursued if of interest, and mehods... it is unclear whether this contribution really adds to the current knowledge... Complete source code, reproducible analyses, and figures are available at https://github.com/lnalborczyk/inner_experience_EMG."
keywords: NHST, Bayesian, fallacy, reanalysis, inner speech, rumination, electromyography
csl: csl/apa7.csl
classoption: man
linenumbers: yes
floatsintext: yes
link-citations: true
numbersections: false
bibliography: [bib/brms_tutorial.bib, bib/r-references.bib]
nocite: | 
  @R-base, @R-knitr, @R-rmarkdown, @R-tidyverse, @R-papaja, @R-wordcountaddin,
  @R-here
output:
  papaja::apa6_pdf:
    keep_tex: true
    highlight: tango
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(wordcountaddin)
library(ggbeeswarm)
library(rmarkdown)
library(tidyverse)
library(patchwork)
library(bayesplot)
library(ggExtra)
library(papaja)
library(readxl)
library(knitr)
library(BEST)
library(here)
library(glue)

@@

# setting up knitr options
knitr::opts_chunk$set(
  cache = TRUE, eval = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE,
  fig.align = "center", dev = "pdf",
  out.width = "100%", fig.pos = "!htb"
  )

@@

# Some thoughts on the data

```{r importing_data}
# importing data
df <- read_excel("data/RUM_master_data.xlsx")

# data reshaping (from wide to long)
# NB: 26 participants in the final dataset...

df2 <- df %>%
    select(ID, order = Order, RUM_FRO_MAX_ln:BAS_OOI_MAX_ln) %>%
    na.omit %>%
    # reshapes data from wide to long format
    pivot_longer(
        cols = starts_with(c("RUM", "DIS", "BAS") ),
        names_to = "condition",
        # names_prefix = "wk",
        values_to = "value",
        values_drop_na = TRUE
        ) %>%
    separate(
        col = condition,
        into = c("condition", "muscle"),
        sep = "_", extra = "merge"
        ) %>%
    # extracts muscle name
    mutate(muscle = substr(muscle, 1, 3) ) %>%
    # from long to wide
    pivot_wider(names_from = muscle, values_from = value)
```

```{r babar, fig.cap = "General plot (N = 26)..."}
# position for groups
pd <- position_dodge(0.8)

# plotting the dtaa
df %>%
    select(ID, order = Order, RUM_FRO_MAX_ln:BAS_OOI_MAX_ln) %>%
    na.omit %>%
    # reshapes data from wide to long format
    pivot_longer(
        cols = starts_with(c("RUM", "DIS", "BAS") ),
        names_to = "condition",
        # names_prefix = "wk",
        values_to = "value",
        values_drop_na = TRUE
        ) %>%
    separate(
        col = condition,
        into = c("condition", "muscle"),
        sep = "_", extra = "merge"
        ) %>%
    # mutate(order = factor(order) ) %>%
    # plots the data
    ggplot(aes(x = condition, y = value) ) +
    # violin plots
    geom_violin(
        scale = "count",
        fill = "black",
        alpha = 0.2,
        colour = "white",
        position = pd,
        # adjust = 0.8,
        # draw_quantiles = 0.5,
        show.legend = FALSE
        ) +
    # plotting individual data points
    geom_dotplot(
        stackdir = "center",
        binaxis = "y",
        position = pd,
        # binwidth = 1,
        dotsize = 0.5,
        alpha = 0.3
        ) +
    stat_summary(
        fun.y = mean,
        geom = "line", size = 1,
        # aes(group = interaction(muscle, order) ),
        aes(group = muscle),
        position = pd,
        show.legend = FALSE
        ) +
    # plotting means
    stat_summary(
        fun.y = "mean",
        geom = "point", shape = 16, size = 3,
        position = pd,
        show.legend = TRUE
        ) +
    # plotting confidence intervals
    stat_summary(
        fun.data = mean_cl_normal,
        geom = "errorbar", size = 1, width = 0,
        fun.args = list(mult = 1.96),
        show.legend = FALSE,
        position = pd
        ) +
    # geom_point(alpha = 0.25, size = 0.5) + 
    # alxtheme +
    facet_grid(~muscle, scales = "free") +
    # scale_color_manual (values = colors2) +
    # scale_fill_manual (values = colors2) +
    labs(x = "Condition", y = "EMG log-amplitude") +
    # theme(legend.title = element_blank() ) +
    # forces lower xlim to 0
    # coord_cartesian(xlim = c(0, 1) ) +
    theme_bw(base_size = 12) +
    scale_colour_brewer(palette = "Dark2", direction = 1) +
    scale_fill_brewer(palette = "Dark2", direction = 1)
```

## Concluding on the null from low-powered studies

https://theethicalskeptic.com/2015/08/17/the-four-types-of-null-hypothesis-fallacy/...

```{r}
# importing data
# df <- read_excel("data/RUM_master_data.xlsx")
```

Unfortunately, the same line of reasoning applies for testing the effect of the order, which is even less powered than the test of the main effect of interest, rendering it practically uninformative...

> "In order to test this, a Bayesian paired samples t-test was conducted for the peak log values of muscle activity between the rumination and distraction conditions. This revealed strong evidence in favour of the alternative hypothesis for the FRO muscle (B10 = 18.79), and moderate evidence in favour of the null hypothesis for the OOS (B10 = 0.232) and OOI (B10 = 0.278) muscles, according to current guidelines for interpreting Bayes factors [43]."

## Does everyone?

```{r everyone, fig.cap = "General plot (mean and 95\\% CI)... showing the inter-individual variability..."}
# does everyone plot
df2 %>%
    pivot_longer(cols = FRO:OOI, names_to = "muscle") %>%
    pivot_wider(names_from = condition, values_from = value) %>%
    mutate(ruminator = ifelse(RUM > DIS, 1, 0) ) %>%
    mutate(ruminator = as.factor(ruminator) ) %>%
    pivot_longer(cols = RUM:BAS, names_to = "condition") %>%
    ggplot(aes(x = condition, y = value, colour = ruminator, fill = ruminator) ) +
    # violin plots
    geom_violin(
        scale = "count",
        fill = "black",
        alpha = 0.2,
        colour = "white",
        position = pd,
        # adjust = 0.8,
        # draw_quantiles = 0.5,
        show.legend = FALSE
        ) +
    # plotting individual data points
    # geom_quasirandom(alpha = 0.3) +
    geom_point(size = 2, alpha = 0.8, show.legend = FALSE) +
    geom_line(aes(group = ID), size = 1, alpha = 0.8, show.legend = FALSE) +
    # plitting the by-group mean
    # stat_summary(
    #     fun.y = mean,
    #     geom = "line", size = 2,
    #     aes(group = interaction(muscle, ruminator) ),
    #     position = pd,
    #     show.legend = FALSE
    #     ) +
    # plotting means
    # stat_summary(
    #     fun.y = "mean",
    #     geom = "point", shape = 16, size = 4,
    #     position = pd,
    #     show.legend = TRUE
    #     ) +
    # plotting confidence intervals
    # stat_summary(
    #     fun.data = mean_cl_normal,
    #     geom = "errorbar", size = 2, width = 0,
    #     fun.args = list(mult = 1.96),
    #     show.legend = FALSE,
    #     position = pd
    #     ) +
    # facetting by muscle
    facet_wrap(~muscle) +
    # axis labels
    labs(x = "Condition", y = "EMG log-amplitude") +
    # forces lower xlim to 0
    # coord_cartesian(xlim = c(0, 1) ) +
    theme_bw(base_size = 12) +
    scale_colour_brewer(palette = "Dark2", direction = -1) +
    scale_fill_brewer(palette = "Dark2", direction = -1)
```

Huge inter-individual variability... which leads to the next point, what is the relation between self-reports and EMG?

## Relation between self-report and EMG correlates

# Discussion and conclusions

@@

Reproducible code and figures are available at https://github.com/lnalborczyk/inner_experience_EMG.

### 2020-09-28 reanalysis.tex
minor updates in manuscript and reanalysis script

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Re-analysing the data from Moffatt et al.~(2020): A textbook illustration of the fallacy of the null hypothesis},
  pdflang={en-EN},
  pdfkeywords={NHST, Bayesian, fallacy, reanalysis},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Correspondence concerning this article should be addressed to Ladislas Nalborczyk, GIPSA-lab, CNRS, Univ. Grenoble Alpes, 11 Rue des Mathématiques, 38400 Saint-Martin-d'Hères, France. E-mail: ladislas.nalborczyk@gipsa-lab.fr}
\keywords{NHST, Bayesian, fallacy, reanalysis}
\usepackage{lineno}

@@

\abstract{
Blah blah\ldots{} Complete source code, reproducible analyses, and figures are available at \url{https://osf.io/ntmgv/}.
}

@@

Wordcount (excluding abstract, references, tables, and figures): 29

@@

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

@@

Supplementary materials, data, reproducible code and figures are available at \url{https://osf.io/ntmgv/}.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Re-analysing the data from Moffatt et al.~(2020): A textbook illustration of the fallacy of the null hypothesis},
  pdflang={en-EN},
  pdfkeywords={NHST, Bayesian, fallacy, reanalysis, inner speech, rumination, electromyography},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Correspondence concerning this article should be addressed to Ladislas Nalborczyk, GIPSA-lab, CNRS, Univ. Grenoble Alpes, 11 Rue des Mathématiques, 38400 Saint-Martin-d'Hères, France. E-mail: ladislas.nalborczyk@gipsa-lab.fr}
\keywords{NHST, Bayesian, fallacy, reanalysis, inner speech, rumination, electromyography}
\usepackage{lineno}

@@

\abstract{
Moffatt et al.~(2020) reported the results of an experiment (N = 26 in the final sample) comparing the facial (surface) electromyographic correlates of induced rumination and distraction, following a stressful event. Based on the observation\ldots{} We show ou this conclusion seems at best hasty\ldots{} Indeed, concluding on the null, relation between self-reports and physiological measures\ldots{} Although the question Moffatt et al.~(2020) pursued if of interest, and mehods\ldots{} it is unclear whether this contribution really adds to the current knowledge\ldots{} Complete source code, reproducible analyses, and figures are available at \url{https://github.com/lnalborczyk/inner_experience_EMG}.
}

@@

Wordcount (excluding abstract, references, tables, and figures): 107

@@

\hypertarget{some-thoughts-on-the-data}{%
\section{Some thoughts on the data}\label{some-thoughts-on-the-data}}

\begin{figure}[!htb]

{\centering \includegraphics[width=1\linewidth]{reanalysis_files/figure-latex/babar-1}

}

\caption{General plot (N = 26)...}\label{fig:babar}
\end{figure}

\hypertarget{concluding-on-the-null-from-low-powered-studies}{%
\subsection{Concluding on the null from low-powered studies}\label{concluding-on-the-null-from-low-powered-studies}}

\url{https://theethicalskeptic.com/2015/08/17/the-four-types-of-null-hypothesis-fallacy/}\ldots{}

Unfortunately, the same line of reasoning applies for testing the effect of the order, which is even less powered than the test of the main effect of interest, rendering it practically uninformative\ldots{}

\hypertarget{does-everyone}{%
\subsection{Does everyone?}\label{does-everyone}}

\begin{figure}[!htb]

{\centering \includegraphics[width=1\linewidth]{reanalysis_files/figure-latex/everyone-1}

}

\caption{General plot (mean and 95\% CI)... showing the inter-individual variability...}\label{fig:everyone}
\end{figure}

Huge inter-individual variability\ldots{} which leads to the next point, what is the relation between self-reports and EMG?

\hypertarget{relation-between-self-report-and-emg-correlates}{%
\subsection{Relation between self-report and EMG correlates}\label{relation-between-self-report-and-emg-correlates}}

\hypertarget{discussion-and-conclusions}{%
\section{Discussion and conclusions}\label{discussion-and-conclusions}}

@@

Reproducible code and figures are available at \url{https://github.com/lnalborczyk/inner_experience_EMG}.
